National Longitudinal Study Reveals Growth in Wellbeing Among Those with Spiritual Beliefs

### 2020-11-16 Manuscript.Rmd
further edits

abstract: |
 Previous research finds an association between spirituality and subjective well-being. However, the widespread use of tautological scales and reliance on cross-sectional data cast doubts. Here, we leverage nine years of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction ($N$ =  20979, New Zealand, 2010-2018). We find that belief in a spirit or life force predicts lower personal wellbeing and life-satisfaction. However, in support of previous theories, beliefs in a spirit or life force predict increasing personal well-being and life satisfaction over time relative to disbelief. These finding holds after adjusting for known demographic influences, and intriguingly, among those who believe in a God but disbelieve in a spirit or life force. The recent growth in spiritual beliefs and decline in traditional religion across many industrial societies motivates futher causal nvestigations of the social and psychological mechanisms through which spiritual beliefs lead to growth in subjective wellbeing. 
  <!-- https://tinyurl.com/ybremelq -->

@@

The relationship between spirituality and psychological well-being is a matter of enduring fascination [@Ellsworth2010-yu]. Previous attempts to quantify the relationship between these two domains reveal a positive association: spirituality predicts well-being [@Russinova2002-rq; @Ginsburg1995-jr; @Zaza2005-ac; @Koenig2010-gk; @Smith2003-re; @Smith2003-re; @Koenig2001-ow]. However, critics raise three challenges.

@@

First, previous research has frequently operationalised spirituality using measures of well-being. For example, The Daily Spiritual Experience Scale, which is widely used in spiritual mental health research, includes items such as "I feel thankful for my blessings" and "I feel deep inner peace or harmony" to measure spirituality [@Underwood2002-hg]. As Koenig observes, it is unsurprising that spirituality predict well-being when the same items are used to measure both constructs. Koenig aptly describes this features as a "tautology" in spirituality/mental health research [@Koenig2008-lv]. Indeed, recent systematic-review of spirituality/mental health research found that nearly 45% of previous studies employe tautological scales [@Garssen2016-kb]. The first challenge, then, is to define measures of spirituality that are not well-being measures reclocked in other words.

A second challenge is to disentangle concepts of spirituality from concepts of religion. The term "spirituality" was first used to describe the religious practices of ascetics and monks, however, more recently, the term has taken on a broader family of meanings [@Swinton2001-vr].  However previous research does not converge on a unique set of meanings. Most studies of spirituality and religion have been conducted among (cross-sectional) North American samples among people who are affiliate with traditional religion [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re;  @Visser2010-kq; @Yonker2012-zg]. However, when spirituality has been operationalised as distinct from religion, some researchers have found a negative association within subjective well-being [@King2013-cg]. Thus, in addition to disentangling the concept of spirituality from concepts of subjective well-being, it is also important to disentangel the concept of spirituality from concepts of religion.

A third challenge is to investigate the relationship between spirituality and subjective well-being over the life-span in nationally diverse samples. To assess whether spiritual beliefs lead to growth, stability, or decline in psychological well-being requires comparing requires repeatedly measuring individuals who differ in spirituality, demography, and religion as these people age, and investigating how their perceive their lives to be turning out [@Garssen2016-km].  However, previous investigations of spirituality and wellbeing, including high-quality reviews and meta-analyses, have relied on cross-sectional samples [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re; @Visser2010-kq; @Yonker2012-zg]. To address the key questions in spirituality psychological health research, then, requires longitudional investigations.

Here, we addresses the three challenges of previous inthe following ways: First, we avoid a tautological elision of "spirituality" and "well-being" by adopting a non-affective measure of spirituality as a believe in a Spirit or Life Force. To  measure subjective well-being we use a well-validated scale for Personal Well-being and a well-validated scale for Life-Satisfaction (see Method). This design avoids measuring spirituality with an indicator of spirituality that is cloacked as psychological well-being, thereby avoiding tautology.  Second, we adopt a four-level categorical indicator for belief-states to clarify differences among people with spiritual, religious, and skeptical beliefs.  We construct this four-level categorical indicator by combining responses to the Spiritual Belief measure with responses to the question: "Do you believe in some form of a spirit or life force?" The levels of our four-level categorical indicator of spiritual and religious beliefs are as follows: (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who nevertheless believe in a Spirit or Life Force (henceforth, SpiritExcludeGod); (3) Those who believe in a God but do not believe in a Spirit or Life force (henceforth, GodAndSpirit); (4) Those who who believe in a both God and a Spirit or Life Force, (henceforth GodExcludesSpirit). Narrowing the concept of spirituality to belief state comes at a price. Such an operationalisation, clearly, cannot simultaneously address all interests in previous research. However, this cost to scope is paid for with the benefits of precision. The spiritual indicators measure cognitive states in which distinct varieties of skeptical beliefs, traditional religious beliefs, and non-traditional spirit beliefs can are disentangled. Third, to assess whether spirit beliefs lead to growth, stability, or decline in psychological well-being, we leverage longitudinal responses from a nationally diverse probability sample year over an nine--year period with nine measurement points (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).

@@

***Belief in a Spirit or a God***  We assess belief in a spirit or a God by asking two non-affective questions: "Do you believe in a spirit or life force" and "Do you believe in a god?" Responses were coded as "Belief in a spirit that excludes a god" for those who indicated a belief in both a spirit or life force and a god, "Belief that includes a god" for those who indicated a belief in a god with or without an accompanying belief in a spirit or life force, and "skeptics" for those who indicated belief in neither a spirit or life force nor a god. These items appear in the NZAVS questionnaire and were developed for the 2005 Eurobarometer. For detailed information pertaining to beliefs, see Table 1.

@@

Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05). We analyzed data from participants who responded to the NZAVS survey at least three four times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of  21,705 (N = 20,979 with complete responses)N = 15,040.  To model predict growth in levels of self-reported psychological subjective well-being among these participants we use generalized linear mixed models. In our models, we account for the non-linear effects of time within participants by including its interaction with both of our predictor variables: "Belief in a spirit that excludes god" and "Belief that includes a god." To handle the dependencies and heterogeneity introduced from the repeated measures, we We also included individual ID as an effect modelled as random-effect. To improve interpretability of our models, we centered and scaled the education, deprivation, and political orientation covariates. Additionally, age was centred at and put into decade-long intervals. 
The model equation(s) are described in equation 1.

\[
\begin{aligned}
\operatorname{{\small LIFESAT/PWI}}  &\sim N \left(\mu,\sigma^2 \right) \\ \mu &=\alpha_{j[i]} + \beta_{1}(\operatorname{Years}) + \beta_{2}(\operatorname{Beliefs}_{\operatorname{\_SpiritExcludesGod\_}})\ + \\
&\quad \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}}) + \beta_{4}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) + \beta_{5}(\operatorname{Age.10yrs.C})\ + \\
&\quad \beta_{6}(\operatorname{Deprivation.S}) + \beta_{7}(\operatorname{Edu.S}) + \beta_{8}(\operatorname{Employed})\ + \\
&\quad \beta_{9}(\operatorname{EthnicCats}_{\operatorname{Maori}}) + \beta_{10}(\operatorname{EthnicCats}_{\operatorname{Pacific}}) + \beta_{11}(\operatorname{EthnicCats}_{\operatorname{Asian}})\ + \\
&\quad \beta_{12}(\operatorname{Male}_{\operatorname{1}}) + \beta_{13}(\operatorname{Partner}_{\operatorname{Has\ }}) + \beta_{14}(\operatorname{Pol.Orient.S})\ + \\
&\quad \beta_{15}(\operatorname{Urban}) + \beta_{16}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{\_SpiritExcludesGod\_}}) + \beta_{17}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\
&\quad \beta_{18}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ \alpha_{j} &\sim N \left(\mu_{\alpha_{j}},\sigma^2_{\alpha_{j}} \right) , \operatorname{ for  Id } j  = 1, \dots ,  J
\end{aligned}
\]

@@

All code for this analysis and a full report of the missing data imputation can be found at https://github.com/jacopastorius/NZAVS_Highland_SpiritHealth2.Table 1. Descriptive Statistics for Theoretical Variables in Each Wave. Belief in a god and belief in a spirit or life force were not measured during Wave 6 (2014). "BXG" indicates a belief in a spirit that excludes a god. "BIG" indicates a belief that includes a god. Many individuals appear in multiple waves.

Table 2. Descriptive Statistics for Demographic Variables in Each Wave. Education level was not measured during Waves 2 (2010) and 3 (2011). Many individuals appear in multiple waves.

@@

Figure 2 about here:

```{r, cache =TRUE, include =FALSE}
# model 2
mod.1 <-lmer (PWI ~  1 +  Years * Beliefs +  Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben, na.action = na.omit)
```

```{r, cache=TRUE, include=FALSE}
# model 1
mod.1 <-lmer (LIFESAT ~ 1 + Years * Beliefs+ Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben, na.action = na.omit)
summary(mod.1)
```

@@

@@

\caption{"Regression Models"}\label{}

@@

 & Model 1 & Model 2 \\

@@

```{r fig.cap = "Combination Coefficient Plots for Personal Well-being and Life Satisfaction Spirit Belief Longitudinal Models"}
knitr::include_graphics("Figs/USEPWI_expected-1.png")
```
  - The effect of Years X Beliefs_SpiritExcludesGod_ is reliably positive  (beta = 0.02, SE = 4.86e-03, std. beta = 0.02, p < .01). We infer that, unlike the skeptic population, people with beliefs in a spirit or life force who do not believe in a God tend to grow in their personal well-beingwellbeing over time.
  - The effect of Years X BeliefsGodAndSpirit is reliably positive  (beta = 0.02, SE = 4.41e-03, std. beta = 0.03, p < .001). We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their personal well-beingwellbeing over time
  - The effect of Years X BeliefsGodExcludesSpirit is not reliable(beta = 1.96e-03, SE = 8.29e-03, std. beta = 2.36e-03, p = 0.813).  Although people who believe in a God but do not believe in a spirit or Life Force tend to have greater personal well-beingwellbeing than those who do not believe in a spirit or life force  (and greater life satisfaction than Skeptics)  this population is not expected to grow in Life Satisfaction over time.

@@

```{r fig.cap = "Prediction Plot Life Satisfaction Spirit Belief Longitudinal Models"}
knitr::include_graphics("Figs/USElifsat_expected-1.png")
```
  - The interaction of Years X Beliefs_SpiritExcludesGod_ is reliably positive (beta = 0.02, SE = 3.66e-03, std. beta = 0.03, p < .001). We infer that, unlike the Skeptic population, people with beliefs in a Spirit or Life Force who do not believe in a God tend to grow in their Life Satisfaction over time.
  - The interaction of Years X BeliefsGodAndSpirit is reliably positive (beta = 0.01, SE = 3.32e-03, std. beta = 0.02, p < .001). We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their Life Satisfaction over time
  - The effect of Years X BeliefsGodExcludesSpirit is not reliable (beta = -1.59e-03, SE = 6.25e-03, std. beta = -2.70e-03, p = 0.799).  Although people who believe in a God but do not believe in a spirit or Life Force tend to have greater life satisfaction than those who do not believe in a Spirit or life force  (and greater life satisfaction than Skeptics)  this population is not expected to grow in Life Satisfaction over time.

@@

Figure 2. Subjective Well-being Predicted Probabilities for Belief in a Spirit or Life Force Over Time in a Listwise Deleted Dataset.

Table 3. Subjective Well-being Coefficient Table for Belief in a Spirit or Life Force. Age was put into units of 10 years and centered at its mean. Education, Deprivation, and Political Orientation were centered at their means and standardized.

@@

The purpose of this study has been to clarify the relationship between belief in a spirit/life force and psychological well-being. Spirituality as a distinct category, separate from religion, is a little-studied phenomenon in the psychology of religion literature.

These findings demonstrate the importance ofinvestigating spiritual beliefs as a domain of interest in its own right. In many Western countries, people are increasingly identifying as "spiritual but not religious" [@Zinnbauer2005-vz; @Swinton2001-vr]. Other studies find that those who have a spiritual understanding of life without a religious framework are at an increased risk of having mental disorders [@King2013-cg]. As such, it is important to understand the ramifications of belief in a spirit for psychological health. 
To assess the relationship between a belief in a spirit or life force and subjective well-being, we improve upon past methods in three ways.

Our findings support critiques of the reliability of past methods.  Still, further research is needed to determine if other aspects of religion and spirituality, such as practices and rituals, might yield different results.

Despite the opposition of our findings to those of many previous studies, there is room for agreement. We show that over time both "BXG" and "BIG" are associated with marginally better subjective well-being. As seen in Figure 2., both "BXG" and "BIG" predict marginally higher personal well-being and life satisfaction as time progresses. So, while those who adhere to these beliefs may start out below their skeptic counterparts in terms of their subjective well-being, they do eventually catch up. We also find that their life satisfaction is actually improving at a greater rate than skeptics. Thus, in the long term, both sustained "BXG" and "BIG" may actually be associated with better subjective well-being.
Our results also help to clarify the relationship among religion, spirituality, and subjective well-being outcomes. BXG and BIG performed almost identically in terms of their ability to predict subjective well-being outcomes. This trend suggests that any kind of existential belief - regardless of whether a god is involved - might affect subjective well-being in much the same way. This result is good news as increasing numbers of individuals continue to report spiritual but not necessarily religious beliefs.

>>>>

abstract: |
 Previous research finds an association between spirituality and subjective well-being. However, the widespread use of tautological spirituality scales, poorly defined concepts of spirituality, and reliance on cross-sectional samples cast doubts. Here, we leverage nine years of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction ($N$ = 20979, New Zealand, 2010-2018). Contrary to previous research, we find that belief in a spirit or life force predicts lower personal well-being and life-satisfaction. However, in support of previous speculation, beliefs in a spirit or life force predict increasing personal well-being and life satisfaction over time relative to disbelief. These finding holds after adjusting for known demographic influences, and intriguingly, among those who believe in a God but disbelieve in a spirit or life force. The recent growth in spiritual beliefs and decline in traditional religion across many industrial societies motivates further causal investigations of the social and psychological mechanisms through which spiritual beliefs lead to growth in subjective well-being. 
  <!-- https://tinyurl.com/ybremelq -->

@@

The relationship between spirituality and psychological well-being is a matter of enduring fascination [@Ellsworth2010-yu]. Most previous attempts to quantify the relationship between these two domains find a positive association: spirituality predicts well-being [@Russinova2002-rq; @Ginsburg1995-jr; @Zaza2005-ac; @Koenig2010-gk; @Smith2003-re; @Smith2003-re; @Koenig2001-ow]. However, critics raise three challenges.

@@

First, previous research has frequently operationalised spirituality using measures of well-being. For example, The Daily Spiritual Experience Scale, which is widely used in spiritual mental health research, includes items such as "I feel thankful for my blessings" and "I feel deep inner peace or harmony" to measure spirituality [@Underwood2002-hg]. As Koenig observes, it is unsurprising that spirituality predict well-being when the same items are used to measure both constructs. Koenig aptly describes this features as a "tautology" in spirituality/mental health research [@Koenig2008-lv]. Indeed, recent systematic-review of spirituality/mental health research found that nearly 45% of previous studies employe tautological scales [@Garssen2016-kb]. The first challenge, then, is to define measures of spirituality that are not well-being measures dressed in other words.

A second challenge is to disentangle concepts of spirituality from concepts of religion. The term "spirituality" was first used to describe the religious practices of ascetics and monks, however, more recently, the term has taken on a broader family of meanings [@Swinton2001-vr].  However previous research does not converge on a unique set of meanings. Most studies of spirituality and religion have been conducted among (cross-sectional) North American samples among people who are affiliate with traditional religion [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re;  @Visser2010-kq; @Yonker2012-zg]. However, when spirituality has been operationalised as distinct from religion, some researchers have found a negative association within subjective well-being [@King2013-cg]. Thus, in addition to disentangling the concept of spirituality from concepts of subjective well-being, it is also important to disentangle the concept of spirituality from concepts of religion.

A third challenge is to investigate the relationship between spirituality and subjective well-being over the life-span in nationally diverse samples. To assess whether spiritual beliefs lead to growth, stability, or decline in psychological well-being requires comparing requires repeatedly measuring individuals who differ in spirituality, demography, and religion as these people age, and investigating how their perceive their lives to be turning out [@Garssen2016-km].  However, previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have relied on cross-sectional samples [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re; @Visser2010-kq; @Yonker2012-zg]. To address the key questions in spirituality psychological health research, then, requires longitudinal investigations.

Here, we addresses the three challenges of previous in the following ways: First, to avoid a tautological elision of "spirituality" and "well-being" we adopt a non-affective measure of spirituality as a "believe in a Spirit or Life Force." On the other side, to measure subjective well-being, we use a well-validated scale for Personal Well-being and a well-validated scale for Life-Satisfaction (see Method). Second, to disentangle spiritual beliefs from traditional religious beliefs and skeptical beliefs we adopt a four-level categorical indicator combining the binary indicator of belief in a Spirit or Life Force with the binary indicator "Do you believe in some form of a spirit or life force?" The levels of this four-level categorical indicator for spiritual, religious, and skeptical beliefs are as follows: (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who nevertheless believe in a Spirit or Life Force; (3) Those who believe in a God but do not believe in a Spirit or Life force; (4) Those who who believe in a both God and a Spirit or Life Force. Clearly such an approach cannot simultaneously address all interests in previous spirituality/well-being research. For example we do not combine our cognitive measure of belief with behavioral measures or attitudinal measures. However, this cost to scope is paid for with benefits to precision. The spiritual indicators measure cognitive states in which distinct varieties of skeptical beliefs, traditional religious beliefs, and non-traditional spirit beliefs are straightforwardly disentangled, and measured within people. Third, to assess whether spirit beliefs lead to growth, stability, or decline in psychological well-being, we leverage longitudinal responses from a nationally diverse probability sample year over an nine--year period with nine measurement points (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).

@@

***Belief in a Spirit or a God***  We assess belief in a spirit or a God by asking two non-affective questions: "Do you believe in a spirit or life force" and "Do you believe in a god?" Responses were coded as (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who believe in a Spirit or Life Force; (3) Believers in a God who disbelieve in a Spirit or Life force; (4) Believers in a God and Spirit or Life Force. These believe were developed for the NZAVS from the 2005 Eurobarometer. For detailed information pertaining to beliefs, see Table 1.

@@

Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05). We analyze data from participants who responded to the NZAVS survey at least three four times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of  21,705 (N = 20,979 with complete responses).  To model growth in subjective well-being we use generalized linear mixed models. In our models, we account for the non-linear effects of time within participants by including the interaction of years (a continuous variable) with Beliefs (a four-level categorical indicator). The baseline for this indicator is Skepticism about a Spirit/Life Forces and Skepticism about a God. To handle the dependencies and heterogeneity introduced from the repeated measures, we include individual ID as an effect modelled as random.  We centered and scaled the education, deprivation, and political conservativism. We centred age was centred at and put into decade-long intervals. 
The model equation(s) are described in Equation 1.

**Equation 1**

\[
\begin{aligned}
\operatorname{{\small LIFESAT/PWI}}  &\sim N \left(\mu,\sigma^2 \right) \\ \mu &=\alpha_{j[i]} + \beta_{1}(\operatorname{Years}) + \beta_{2}(\operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\
&\quad \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}}) + \beta_{4}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) + \beta_{5}(\operatorname{Age.10yrs.C})\ + \\
&\quad \beta_{6}(\operatorname{Deprivation.S}) + \beta_{7}(\operatorname{Edu.S}) + \beta_{8}(\operatorname{Employed})\ + \\
&\quad \beta_{9}(\operatorname{EthnicCats}_{\operatorname{Maori}}) + \beta_{10}(\operatorname{EthnicCats}_{\operatorname{Pacific}}) + \beta_{11}(\operatorname{EthnicCats}_{\operatorname{Asian}})\ + \\
&\quad \beta_{12}(\operatorname{Male}_{\operatorname{1}}) + \beta_{13}(\operatorname{Partner}_{\operatorname{Has\ }}) + \beta_{14}(\operatorname{Pol.Orient.S})\ + \\
&\quad \beta_{15}(\operatorname{Urban}) + \beta_{16}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}}) + \beta_{17}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\
&\quad \beta_{18}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ \alpha_{j} &\sim N \left(\mu_{\alpha_{j}},\sigma^2_{\alpha_{j}} \right) , \operatorname{ for  Id }~j~= 1, \dots~J = 20,979~\operatorname{Ids}
\end{aligned}
\]

@@

All code for this analysis and a full report of the missing data imputation can be found at https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing.

@@

```{r, cache =TRUE, include =FALSE}
# model 2
mod.1 <-lmer (PWI ~  1 +  Years * Beliefs +  Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben, na.action = na.omit)
```
```{r, cache=TRUE, include=FALSE}
# model 1
mod.1 <-lmer (LIFESAT ~ 1 + Years * Beliefs+ Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben, na.action = na.omit)
summary(mod.1)
```

@@

Figure 1 about here:

@@

\caption{"Longitudinal Models"}\label{}

@@

 & Personal Well-Being & Life Satisfaction \\

@@

```{r fig.cap = "Combination Coefficient Plots for Personal Well-being and Life Satisfaction Spirit Belief Longitudinal Models"}
knitr::include_graphics("Figs/USEPWI_expected-1.png")
```
Focusing on the interaction of time with belief-type on personal well-being we observe:

  - The effect of Years X SpiritExcludesGod_ is reliably positive  (beta = 0.02, SE = 4.86e-03, std. beta = 0.02, p < .01). We infer that, unlike the skeptic population, people with beliefs in a spirit or life force who do not believe in a God tend to grow in their personal well-being over time.
  - The effect of Years X GodAndSpirit is reliably positive  (beta = 0.02, SE = 4.41e-03, std. beta = 0.03, p < .001). We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their personal well-being over time
  - The effect of Years X GodExcludesSpirit is not reliable(beta = 1.96e-03, SE = 8.29e-03, std. beta = 2.36e-03, p = 0.813).  Thus, God believers who doubt in a Spirit or Life Force start maintain higher personal well-being than do other groups, but they are not expected to grow in personal well-being over time.

@@

```{r fig.cap = "Prediction Plot Life Satisfaction Spirit Belief Longitudinal Models"}
knitr::include_graphics("Figs/USElifsat_expected-1.png")
```
  - The interaction of Years X Beliefs_SpiritExcludesGod_ is reliably positive (beta = 0.02, SE = 3.66e-03, std. beta = 0.03, p < .001). We infer that, unlike the Skeptic population, people with beliefs in a Spirit or Life Force who do not believe in a God tend to grow in their Life Satisfaction over time.
  - The interaction of Years X BeliefsGodAndSpirit is reliably positive (beta = 0.01, SE = 3.32e-03, std. beta = 0.02, p < .001). We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their Life Satisfaction over time
  - The effect of Years X BeliefsGodExcludesSpirit is not reliable (beta = -1.59e-03, SE = 6.25e-03, std. beta = -2.70e-03, p = 0.799).  Thus, God believers who doubt in a Spirit or Life Force start maintain higher Life-Satisfaction than do other groups, but they are not expected to grow in Life-Satisfaction over time.

@@

## Imputed data analysis

To adjust for potential biases from missing responses, we conducted a supplementary analysis in which we imputed missing values over the the longitudinal sample. This analysis is describe in Supplement 1. Averaging over the missing responses does not change inference, however we observe growth across all four groups in both personal well-being and life satisfaction. Additionally we observe substantially stronger effects for the slope of Spirit Beliefs on both personal well-being and life-satisfaction. Figures 3 and Figures 4 describe the predicted slopes of change. The inference from this analysis remains the same as from the analysis in which we delete missing values row-wise, by panel wave.  In all analyses those who believe in a spirit or life-force are expected to experience greater growth in well-being over time.

Figure 3 about here

Figure 4 about here

@@

The purpose of this study has been to clarify the relationship between belief in a spirit/life force and subjective well-being, addressing three challenges from previous research: (1) tautological measures of spirituality and subjective well-being; (2) insufficiently distinct measures of spirituality and religion; (3) cross-sectional samples. Here, we ask whether beliefs in a Spirit or Life Force predict growth in personal well-being and life-satisfaction over eight years in populations who differ in the this belief, and who also differ in their belief in a God.  We find that

These findings demonstrate the importance of investigating spiritual beliefs as a domain of interest in its own right. In many Western countries, people are increasingly identifying as "spiritual but not religious" [@Zinnbauer2005-vz; @Swinton2001-vr]. Other studies find that those who have a spiritual understanding of life without a religious framework are at an increased risk of having mental disorders [@King2013-cg]. As such, it is important to understand the ramifications of belief in a spirit for psychological health. 
To assess the relationship between a belief in a spirit or life force and subjective well-being, we improve upon past methods in three ways.

Despite the opposition of our findings to those of many previous studies, there is room for agreement. We show that over time both "BXG" and "BIG" are associated with marginally better subjective well-being. As seen in Figure 2., both "BXG" and "BIG" predict marginally higher personal well-being and life satisfaction as time progresses. So, while those who adhere to these beliefs may start out below their skeptic counterparts in terms of their subjective well-being, they do eventually catch up. We also find that their life satisfaction is actually improving at a greater rate than skeptics. Thus, in the long term, both sustained "BXG" and "BIG" may actually be associated with better subjective well-being.

Our results also help to clarify the relationship among religion, spirituality, and subjective well-being outcomes. BXG and BIG performed almost identically in terms of their ability to predict subjective well-being outcomes. This trend suggests that any kind of existential belief - regardless of whether a god is involved - might affect subjective well-being in much the same way. This result is good news as increasing numbers of individuals continue to report spiritual but not necessarily religious beliefs.
